A longitudinal analysis of the privacy paradox

### 2021-05-26 manuscript.RmD
- last minor edits after checking proofs

abstract: |
  The privacy paradox states that people’s concerns about online privacy are unrelated to their online sharing of personal information. On the basis of a representative sample of the German population, which includes `r knitr::load_cache(label = "load-data", object = "n_complete")` respondents interviewed at three waves separated by 6 months, we investigate the privacy paradox from a longitudinal perspective. Using a cross-lagged panel model with random intercepts, we differentiate between-person relations from within-person effects. Results revealed that people who were more concerned about their online privacy than others also shared slightly less personal information and had substantially more negative attitudes toward information sharing (between-person level). People who were more concerned than usual also shared slightly less information than usual (within-person level). We found no long-term effects of privacy concerns on information sharing or attitudes 6 months later. The results provide further evidence against the privacy paradox, but more research is needed to better understand potential causal relations.
keywords          : "privacy paradox, privacy concerns, information sharing, longitudinal analysis, structural equation modeling"
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
class             : "man"
mask              : no
# to compile as pdf
# (a) move file "_site.yml" to folder "archive"
# (b) uncomment the following lines until incl. "\raggedbottom"
# (c) comment lines after "output: html_document ..."
output            : papaja::apa6_pdf
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
# output:
#   html_document:
#     toc: TRUE
#     toc_float:
#       collapsed: FALSE
#     toc_depth: 2
bibliography      : 
  - [bibliography/references.bib]
  - [bibliography/r-references.bib]
csl               : "sage-harvard.csl"
---

@@

However, current research on the privacy paradox has one major limitation. 
To the best of our knowledge, most empirical studies conducted so far have investigated the privacy paradox from a between-person perspective.
By employing empirical tests of relations between people (e.g., cross-sectional questionnaires analyzed with multiple regression or Pearson correlations), studies have analyzed whether people who are more concerned than _others_ also share less personal information than _others_.
Although such a perspective is interesting and represents a viable first step, it cannot make informed claims regarding causality.
The privacy paradox, however, implies a causal perspective:
Does a person, if _he or she_ becomes more concerned about online privacy, then also share less personal information?
This mismatch is problematic because although between-person relations are, except for some special cases, a _necessary_ condition for causal within-person effects, they are by no means a _sufficient_ one.
For example, it could be that the between-person relation is determined by other third variables. 
Hence, as the next step in investigating the privacy paradox and to better understand the causal relation between privacy concerns and information sharing, it is necessary to conduct studies with within-person designs.

@@

@acquistiLossesGainsHyperbolic2003 were among the first to argue that the online disclosure of personal information is paradoxical. 
"Experiments reveal that very few individuals actually take any action to protect their personal information, even when doing so involves limited costs" (p.1). 
Three years later, @barnesPrivacyParadoxSocial2006 discussed the behavior of young people on SNSs, popularizing the term _privacy paradox_. 
Barnes listed six aspects of online behavior that she considered to be particularly paradoxical: (a) illusion of privacy, (b) high quantity of information sharing, (c) attitude behavior discrepancy, (d) lack of privacy concerns, (e) lack of privacy literacy, and (f) fabrication of false information.
@norbergPrivacyParadoxPersonal2007 were one of the first to empirically analyze the privacy paradox explicitly.
The study found a mismatch between concerns and behavior, which is aligned with several other experimental studies conducted at the time [@beresfordUnwillingnessPayPrivacy2012; @hannOvercomingOnlineInformation2007; @hubermanValuatingPrivacy2005].

@@

Several dimensions of privacy have been proposed.
For example, it is possible to distinguish a vertical and a horizontal level [@masurSituationalPrivacySelfdisclosure2018].
Whereas the vertical level captures privacy from authorities, institutions, or companies, horizontal privacy addresses privacy from peers, colleagues, or other people.
When it comes to concerns in general, interestingly they do not seem to be established as a stand-alone theoretical concept in psychology [@colmanDictionaryPsychology2015].
Concerns are usually understood as an uneasy mix of "interest, uncertainty, and apprehension" [@merriam-websterDefinitionConcern2018].
As a theoretical construct, privacy concerns can hence be categorized as an affective motivational disposition.
Taken together, concerns about online privacy represent how much an individual is motivated to focus on his or her control over a voluntary withdrawal from other people or societal institutions on the Internet, accompanied by an uneasy feeling that his or her privacy might be threatened.

The online sharing of personal information, on the other hand, captures how much person-related information people share when they use the Internet.
Information sharing can be differentiated from communication and self-disclosure.
Communication is broad, because it comprises all verbal and nonverbal information that is emitted [e.g., @watzlawickPragmaticsHumanCommunication2011].
Self-disclosure is more narrow, because it focuses on deliberate revelations about the true self to others, which including aspects such as personal fears, values, or plans [e.g., @jourardTransparentSelf1964].
Information sharing is even more specific, because it addresses only person-related information,  including information about their age, sex, name, address, health, and finances.

@@

Specifically, two of the most influential factors that affect online information sharing are (a) strong subjective norms [@heirmanPredictingAdolescentsDisclosure2013] and (b) expected benefits [@krasnovaOnlineSocialNetworks2010]. 
In other words, users often prioritize social support, special offers, or improved services, accepting that their privacy will be diminished.
Sometimes, privacy concerns do not relate information sharing, because users lack the skills, knowledge, or literacy to change their online behavior, creating feelings of apathy or cynicism [@hargittaiWhatCanReally2016; @hoffmannPrivacyCynicismNew2016].
Likewise, personal information is also often shared by others, a phenomenon described as "networked privacy" [@marwickNetworkedPrivacyHow2014], which further reduces the power of individuals to determine how much personal information can be found online.
@trepteRiskyBehaviorsHow2014 listed several factors that can additionally attenuate the relation:
lack of strength of concerns,
absence of negative personal experiences, 
or situational constraints due to social desirability.
In conclusion, also in the context of the privacy paradox it is _not_ reasonable to expect a _perfect_ relation between attitudes and behaviors.
However, we should still expect to find a relation that is _small_ or _moderate_.

There are also some methodological explanations as to why some studies did not detect statistically significant relations.
Researchers are always confronted with the so-called _Duhem-Quine problem_, according to which it is impossible to test theories in isolation, because empirical tests always rely on auxiliary assumptions [@dienesUnderstandingPsychologyScience2008]. 
In other words, if a psychological experiment fails, we do not know whether the theory is wrong or the questionnaire subpar. 
This tenet is particularly relevant for the privacy paradox: 
Detecting statistical significance for small effects---and, again, we should expect to find small effects---is more challenging because it means that large samples are necessary to guarantee sufficient statistical power.[^power]
Precisely, in order to be capable of detecting a correlation between privacy concerns and information sharing in 95% of all cases, which @baruhOnlinePrivacyConcerns2017 estimated to be _r_ = -.13, we need a sample of _N_ = `r pwr.r.test(r = -.13, sig.level = .05, power = .95)$n %>%  round(0)` people. 
The reality, however, looks different: 
In their meta-analysis, @baruhOnlinePrivacyConcerns2017 reported a median sample size of _N_ = 300, which can explain why several studies did not find significant effects.

@@

[^1]: To explain, with Hypothesis 1, we compare _different_ people with one another by analyzing their _average_ values across all measurements.
In other words, does a person, who is generally more concerned than others, also generally share less information than others?
With Hypothesis 2, we compare _specific_ measurements within the _same_ person.
In other words, does a person, if he/she is more concerned on T1 than on average, share more or less information on T1 than on average?

@@

Although short-term effects are likely, it is still unclear whether long-term effects exist as well.
First, when analyzing potential long-term effects, it is important to choose an interval that is both plausible and relevant.
(It makes a large difference whether the effects of alcohol consumption on driving performance are tested after say 1 minute, 1 hour, or 1 day.) 
One factor that determines an interval's optimal length is the stability of the variables [@dormannOptimalTimeLags2015].
Privacy concerns and privacy attitudes are predominantly trait-like constructs with high stabilities, which is why they necessitate longer intervals.
Other studies with comparable research questions have therefore used an interval of 6 months [e.g., @valkenburgEffectsInstantMessaging2009], which we adopt also in this study.

@@

This study is part of a large-scale project which investigates the development of privacy and self-disclosure, including several other variables. 
Other publications linked to the project can be accessed at [link blinded during review].
The data come from a longitudinal paper-and-pencil questionnaire study, in which a representative sample of the German population (16 years and older) was surveyed on overall five occasions.
The data can be downloaded from [link blinded during review].

@@

We tested the factorial validity of all measures using confirmatory factor analysis (CFA). 
Each CFA included the items from all three waves. 
For each item, factor loadings were constrained to be equal across waves.
Constrained and unconstrained models were compared using $\chi^2$ differences tests.
All results were nonsignificant, suggesting longitudinal factorial invariance.
The measures showed good composite reliability in all three waves. 
Graphical displays of the variables' distributions showed that privacy concerns were skewed to the left, privacy attitudes were normally distributed, and information sharing was skewed to the right (Figure \@ref(fig:Distributions), diagonal).
We calculated intra-class correlation coefficients to quantify how much variance in the variables' factor scores could be attributed to between-person differences.
An English translation of the original German items can be found in the OSM.

@@

```{r RICLPM, fig.cap="Visual representation of the estimated random-intercept cross-lagged panel model (RI-CLPM).", fig.align="center"}
knitr::include_graphics("figures/model_7.png")
```

@@

 - Context (e.g., professional, social, commercial, or health-related);
 - Situation (e.g., new, habitualized, or unexpected);
 - Mood (e.g., positive vs. negative);
 - Extent of control (high vs. low);
 - Type of information processing applied (implicit, heuristic, or peripheral vs. explicit, analytic, or central);
 - Existence of bias (e.g., overconfidence, optimism, comparative optimism, or hyperbolic discounting);
 - Type of information (e.g., sensitive vs. superficial, biographic, or person-related);
 - Benefit immediacy and risk diffusion (high vs. low);
 - Object of investigation (e.g., individual people, interactions between people, developmental perspectives, critical incidents, societal structures, or historical developments).

@@

 - Analysis design and perspective (e.g., theoretical, experimental, questionnaire-based, interview-based, ethnographic, or computational);
 - Quality of measurement (high vs. low; low quality less likely to detect statistical significance);
 - Sample size (small vs. large; small samples less likely to detect statistical significance);
 - Statistical analysis (e.g., SEM vs. Regression; analyses without error control less likely to find statistical significance);
 - Operationalization (e.g., concerns vs. risk perceptions vs. behavioral attitudes; the more specific, the stronger the relation).

>>>>

abstract: |
  The privacy paradox states that people’s concerns about online privacy are unrelated to their online sharing of personal information. On the basis of a representative sample of the German population, which includes `r knitr::load_cache(label = "load-data", object = "n_complete")` respondents interviewed at three waves separated by 6 months, we investigate the privacy paradox from a longitudinal perspective. Using a cross-lagged panel model with random intercepts, we differentiate between-person relations from within-person effects. Results revealed that people who were more concerned about their online privacy than others also shared slightly less personal information and had substantially more negative attitudes toward information sharing (between-person level). People who were more concerned than usual also shared slightly less information than usual (within-person level). We found no long-term effects of privacy concerns on information sharing or attitudes 6 months later. The results provide further evidence against the privacy paradox, but more research is needed to better understand potential causal relations.
keywords          : "privacy paradox, privacy concerns, information sharing, longitudinal analysis, structural equation modeling"
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
class             : "man"
mask              : no
# to compile as pdf
# (a) move file "_site.yml" to folder "archive"
# (b) uncomment the following lines until incl. "\raggedbottom"
# (c) comment lines after "output: html_document ..."
# output            : papaja::apa6_pdf
# header-includes:
#   - \setlength{\parskip}{0em}
#   - \raggedbottom
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
bibliography      : 
  - [bibliography/references.bib]
  - [bibliography/r-references.bib]
csl               : "sage-harvard.csl"
---

@@

However, current research on the privacy paradox has one major limitation. 
To the best of our knowledge, most empirical studies conducted so far have investigated the privacy paradox from a between-person perspective.
By employing empirical tests of relations between people (e.g., cross-sectional questionnaires analyzed with multiple regression or Pearson correlations), studies have analyzed whether people who are more concerned than _others_ also share less personal information than _others_.
Although such a perspective is interesting and represents a viable first step, it cannot make informed claims regarding causality.
The privacy paradox, however, implies a causal perspective:
Does a person, if _they_ become more concerned about online privacy, then also share less personal information?
This mismatch is problematic because although between-person relations are, except for some special cases, a _necessary_ condition for causal within-person effects, they are by no means a _sufficient_ one.
For example, it could be that the between-person relation is determined by other third variables. 
Hence, as the next step in investigating the privacy paradox and to better understand the causal relation between privacy concerns and information sharing, it is necessary to conduct studies with within-person designs.

@@

@acquistiLossesGainsHyperbolic2003 were among the first to argue that the online disclosure of personal information is paradoxical. 
"Experiments reveal that very few individuals actually take any action to protect their personal information, even when doing so involves limited costs" (p.1). 
Three years later, @barnesPrivacyParadoxSocial2006 discussed the behavior of young people on SNSs, popularizing the term _privacy paradox_. 
Barnes considered the following six aspects of online behavior particularly paradoxical: (a) illusion of privacy, (b) high quantity of information sharing, (c) attitude behavior discrepancy, (d) lack of privacy concerns, (e) lack of privacy literacy, and (f) fabrication of false information.
@norbergPrivacyParadoxPersonal2007 were one of the first to empirically analyze the privacy paradox explicitly.
The study found a mismatch between concerns and behavior, which is aligned with several other experimental studies conducted at the time [@beresfordUnwillingnessPayPrivacy2012; @hannOvercomingOnlineInformation2007; @hubermanValuatingPrivacy2005].

@@

Several dimensions of privacy have been proposed.
For example, it is possible to distinguish a vertical and a horizontal level [@masurSituationalPrivacySelfdisclosure2018].
Whereas the vertical level captures privacy from authorities, institutions, or companies, horizontal privacy addresses privacy from peers, colleagues, or other people.
When it comes to concerns in general, interestingly they do not seem to be established as a stand-alone theoretical concept in psychology [@colmanDictionaryPsychology2015].
Concerns are usually understood as an uneasy mix of "interest, uncertainty, and apprehension" [@merriam-websterDefinitionConcern2018].
As a theoretical construct, privacy concerns can hence be categorized as an affective motivational disposition.
Taken together, concerns about online privacy represent how much an individual is motivated to focus on their control over a voluntary withdrawal from other people or societal institutions on the Internet, accompanied by an uneasy feeling that their privacy might be threatened.

The online sharing of personal information, on the other hand, captures how much person-related information people share when they use the Internet.
Information sharing can be differentiated from communication and self-disclosure.
Communication is broad, because it comprises all verbal and nonverbal information that is emitted [e.g., @watzlawickPragmaticsHumanCommunication2011].
Self-disclosure is more narrow, because it focuses on deliberate revelations about the true self to others, including aspects such as personal fears, values, or plans [e.g., @jourardTransparentSelf1964].
Information sharing is even more specific, because it addresses only person-related information,  including information about their age, sex, name, address, health, and finances.

@@

Specifically, two of the most influential factors that affect online information sharing are (a) strong subjective norms [@heirmanPredictingAdolescentsDisclosure2013] and (b) expected benefits [@krasnovaOnlineSocialNetworks2010]. 
In other words, users often prioritize social support, special offers, or improved services, accepting that their privacy will be diminished.
Sometimes, privacy concerns do not relate to information sharing, because users lack the skills, knowledge, or literacy to change their online behavior, creating feelings of apathy or cynicism [@hargittaiWhatCanReally2016; @hoffmannPrivacyCynicismNew2016].
Likewise, personal information is also often shared by others, a phenomenon described as "networked privacy" [@marwickNetworkedPrivacyHow2014], which further reduces the power of individuals to determine how much personal information can be found online.
@trepteRiskyBehaviorsHow2014 listed several factors that can additionally attenuate the relation:
lack of strength of concerns,
absence of negative personal experiences, 
or situational constraints due to social desirability.
In conclusion, also in the context of the privacy paradox it is _not_ reasonable to expect a _perfect_ relation between attitudes and behaviors.
However, we should still expect to find a relation that is _small_ or _moderate_.

There are also some methodological explanations as to why some studies did not detect statistically significant relations.
Researchers are always confronted with the so-called _Duhem-Quine problem_, according to which it is impossible to test theories in isolation, because empirical tests always rely on auxiliary assumptions [@dienesUnderstandingPsychologyScience2008]. 
In other words, if a psychological experiment fails, we do not know whether the theory is wrong or the questionnaire subpar. 
This tenet is particularly relevant for the privacy paradox: 
Detecting statistical significance for small effects---and, again, we should expect to find small or moderate effects---is more challenging because it means that large samples are necessary to guarantee sufficient statistical power.[^power]
Precisely, in order to be capable of detecting a correlation between privacy concerns and information sharing in 95% of all cases, which @baruhOnlinePrivacyConcerns2017 estimated to be _r_ = -.13, we need a sample of _N_ = `r pwr.r.test(r = -.13, sig.level = .05, power = .95)$n %>%  round(0)` people. 
The reality, however, looks different: 
In their meta-analysis, @baruhOnlinePrivacyConcerns2017 reported a median sample size of _N_ = 300, which can explain why several studies did not find significant effects.

@@

[^1]: To explain, with Hypothesis 1, we compare _different_ people with one another by analyzing their _average_ values across all measurements.
In other words, does a person, who is generally more concerned than others, also generally share less information than others?
With Hypothesis 2, we compare _specific_ measurements within the _same_ person.
In other words, does a person, if they are more concerned on T1 than on average, share more or less information on T1 than on average?

@@

Although short-term effects are likely, it is still unclear whether long-term effects exist as well.
First, when analyzing potential long-term effects, it is important to choose an interval that is both plausible and relevant.
It makes a large difference whether the effects of alcohol consumption on driving performance are tested after say 1 minute, 1 hour, or 1 day. 
One factor that determines an interval's optimal length is the stability of the variables [@dormannOptimalTimeLags2015].
Privacy concerns and privacy attitudes are predominantly trait-like constructs with high stabilities, which is why they necessitate longer intervals.
Other studies with comparable research questions have therefore used an interval of 6 months [e.g., @valkenburgEffectsInstantMessaging2009], which we adopt also in this study.

@@

This study is part of a large-scale project which investigates the development of privacy and self-disclosure, including several other variables. 
Other publications linked to the project can be accessed at https://osf.io/y35as/.
The data come from a longitudinal paper-and-pencil questionnaire study, in which a representative sample of the German population (16 years and older) was surveyed on overall five occasions.
The data can be downloaded from https://doi.org/10.7802/1937.

@@

We tested the factorial validity of all measures using confirmatory factor analysis (CFA). 
Each CFA included the items from all three waves. 
For each item, factor loadings were constrained to be equal across waves.
Constrained and unconstrained models were compared using $\chi^2$ differences tests.
All results were nonsignificant, suggesting longitudinal factorial invariance.
The measures showed good composite reliability in all three waves. 
Graphical displays of the variables' distributions showed that privacy concerns were skewed to the left, privacy attitudes were normally distributed, and information sharing was skewed to the right (Figure \@ref(fig:Distributions), diagonal).
We calculated intraclass correlation coefficients (ICCs) to quantify how much variance in the variables' factor scores could be attributed to between-person differences.
An English translation of the original German items can be found in the online supplementary material.

@@

```{r RICLPM, fig.cap="The estimated random-intercept cross-lagged panel model (RI-CLPM).", fig.align="center"}
knitr::include_graphics("figures/model_7.png")
```

@@

 - context (e.g., professional, social, commercial, or health-related);
 - situation (e.g., new, habitualized, or unexpected);
 - mood (e.g., positive vs. negative);
 - extent of control (high vs. low);
 - type of information processing applied (implicit, heuristic, or peripheral vs. explicit, analytic, or central);
 - existence of bias (e.g., overconfidence, optimism, comparative optimism, or hyperbolic discounting);
 - type of information (e.g., sensitive vs. superficial, biographic, or person-related);
 - benefit immediacy and risk diffusion (high vs. low);
 - object of investigation (e.g., individual people, interactions between people, developmental perspectives, critical incidents, societal structures, or historical developments).

@@

 - analysis design and perspective (e.g., theoretical, experimental, questionnaire-based, interview-based, ethnographic, or computational);
 - quality of measurement (high vs. low; low quality less likely to detect statistical significance);
 - sample size (small vs. large; small samples less likely to detect statistical significance);
 - statistical analysis (e.g., SEM vs. Regression; analyses without error control less likely to find statistical significance);
 - operationalization (e.g., concerns vs. risk perceptions vs. behavioral attitudes; the more specific, the stronger the relation).
